Thermal reactionomes reveal adaptive responses to thermal extremes in warm

### 2014-10-13 doc/ApTranscriptome_manuscript.Rmd
edits from NG

**Limit 250 words, at 250! If you add words...find others to delete!**

Thermal tolerance is a key trait governing the distributions of species and is useful in predicting their responses to climate change. To quantitatively evaluate among three hypotheses for the evolution of thermal tolerance, we describe a novel analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient. We characterized the thermal reactionomes of two common ant species in the eastern U.S: the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*. We found that at least 8% of genes were thermally responsive. The cool-adapted *A. picea* induced expression of more genes in response to extreme temperatures than *A. carolinensis*, including heat shock proteins, consistent with the *enhanced defense* hypothesis. In contrast, *A. carolinensis* displayed higher constitutive expression of responsive genes and required greater changes towards high, but not low, temperatures to induce changes in gene expression than did *A. picea*. This pattern was consistent with *tolerance* hypothesis at high, but not low, temperatures. We found no evidence for the *genetic assimilation* hypothesis as there was no trade-off between constitutive and inducible levels of gene expression. Overall, our results indicate that the northern *A. picea* actively resists thermal shock by synthesizing heat shock and other protective molecules to maintain function, whereas the southern *A. carolinensis* relies upon other mechanisms to tolerate thermal stress. These results are in line with expectations that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.

@@

**Limit 120 words, at 120! If you add words...find others to delete!**

Coping with progressive changes in climate requires both short-term physiological acclimation and long-term adaptive changes. To better understand the relationship between these two timescales, we compared gene expression as a function of temperature in two closely-related ant species: the cool-climate *Aphaenogaster picea* and the warm-climate *A. carolinensis*. Surprisingly, we found divergent transcriptomic responses in these species. *Aphaenogaster picea* responded to high and low temperatures through upregulation of a larger proportion of its transcriptome with a robust heat shock response. In contrast, *A. carolinensis* required more extreme temperatures to trigger a protective response and downregulated genes related to metabolism. These results suggest that adaptation to chronic high temperatures involves qualitatively different mechanisms than those used to cope with rarely encountered stressors.

@@

Temperature regulates biological activity and shapes diversity from molecular to macroecological scales [@brown2004; @kingsolver2009]. Many species, especially small insects, live at temperatures close to their thermal limits and are at risk from current increases in temperature [@deutsch2008; @kingsolver2013; @sunday2014]. Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, depends on a complex interplay between the structural integrity of cellular components and activiation of physiological response mechanisms to prevent and/or repair damage [@huey1989; @richter2010]. Thermal defense strategies are shaped by the environmental regime organisms experience [@angilletta_jr2003] and thermal limits vary considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011]. Outside of candidate genes [@krebs1998], little is known about the evolution of thermal tolerance or the link between short-term physiological acclimation and longer-term adaptation to novel temperature regimes. This information is critical for understanding the adaptive potential of species to future climates [@hoffmann2008].

To address this gap of knowlege, we need information on the extent to which selection has acted upon the diversity genes and their plasticity of expression involved in thermal tolerance [@hoffmann2008; @somero2011]. In recent years, whole organism gene expression approaches (e.g. transcriptomics) using high-throughput RNA sequencing (RNAseq) technology have been widely applied to identify genes involved in thermal tolerance [@meyer2011; @teets2012; @barshis2013; @oneil2014] and other traits. Such studies typically use an ANOVA-type experimental or sampling design, with only a few environmental levels, and often find only a few dozen to hundred genes with differential expression in different thermal regimes. However, temperature and other environmental factors vary continuously in nature. As a result, such comparisons (e.g. high vs. low temperatures) are likely to miss key differences that are due not just to whether it is hot, but rather how hot it is. Continuous variation is better characterized with a reaction norm approach, which describes variation in the phenotype of a single genotype across an environmental gradient [@gomulkiewicz1992]. Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014], and differences in the shape of reaction norms are often greater than differences in mean values at both the species and population levels [@murren2014].

Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient. We use this method to identify genes that are thermally responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* (part of the *A. rudis*-species complex, B. DeMarco personal communication) and *A. picea* [@umphrey1996]. *Aphaenogaster* are some of the most common ant in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. *Aphaenogaster carolinensis* (mean annual temperature (MAT) at collection site = 14.6°C, seasonality = 7,678) experiences significantly warmer and less seasonally variable temperatures than *A. picea* (MAT = 4.6°C, seasonality = 10,008; [@hijmans2005]), with correspondingly higher critical maximum temperatures (44.7°C for *A. carolinensis* versus 41.4°C for *A. picea*; Supplemental Material) and minimum temperatures (6.1°C for *A. carolinensis* versus -0.1°C for *A. picea*; Supplemental Material) allowing us to investigate adaptation to both lower and upper thermal extremes in this system.

@@

To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) using Illumina short read sequencing technology, generated the *de novo* transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis (Supplemental Material). We used these data to quantitatively test three mechanistic hypotheses of thermal adaptation. First, the *enhanced defense* hypothesis [@hofmann1996; @feder1999; @kultz2005] proposes that the stress-adapted species extend their thermal limits through a stronger induced response to provide greater protection from damage. This hypothesis would predict that *A. picea* induces more genes in response to low temperatures, whereas *A. carolinensis* induces more genes in response to high temperatures. Additionally, upregulation of molecular pathways shared by both species should be stronger or occur in response to less initial temperatures stress for more commonly-experienced thermal extremes; low temperatures for *A. picea*, high temperatures for *A. carolinensis*. Second, the *genetic assimiliation* hypothesis [@waddington1953; @sikkink2014] proposes that exposure to more frequent stressors selects for a shift from inducible to constitutive expression of stress-response genes. This hypothesis predicts that transcripts responsive to high temperatures in *A. picea* will have higher constitutive expression in *A. carolinensis*, while transcripts responsive to low temperatures in *A. carolinensis* will have higher constitutive expression in *A. picea*. Finally, the *tolerance* hypothesis [@cowles1939; @fields2001] proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural [@lockwood2012] or behavioral [@gordon2013; @sunday2014] adapations. This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways. Of course, these hypotheses are not mutually exclusive. By taking a reactionome approach, we can quantify the relative importance of each of these mechanisms to thermal adapatation as the proportion of the transcriptome matching each of these predicted patterns.

@@

The combined *Aphaenogaster* transcriptome assembly contained 98,186 transcripts. About half of these (51,246) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae. Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts, or 8% of the total, with expression patterns that were thermally responsive at a 5% False Discovery Rate [@benjamini1995]. Of these, `r prettyNum(length(quadratic.lms), big.mark = ",")` had a non-linear relationship with temperature that would likely have been missed with a standard differential expression approach. The result is the same if we focus only on those transcript with a BLAST hit (4,618 transcripts out of 51,246). However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] suggesting that this estimate is likely a lower-bound for the true proportion of genes that are thermally responsive.

```{r type_table, echo=FALSE, results='asis', fig.cap = "Table of the number of thermally-responsive transcripts by expression type for *A. carolinensis* and *A. picea*. **Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are thermally responsive in one species but not the other."}

@@

The total number of thermally-responsive transcripts did not differ between species (...stat...). Within this set, there was significant variation in response both among transcripts and between the two species. Transcripts varied both qualitatively and quantitatively in their their thermal reaction norms. We grouped the thermally-responsive transcripts for each species into five expression categories based on their reaction norms: transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (**Not Responsive**). For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C). The majority of the responsive genes were in the **Low** category for both species (42% and 54% for *A. carolinensis* and *A. picea*, respectively), followed by **Intermediate** (29%), **High** (12%),**Bimodal** (10%) and **Not Responsive** (8%) for *A. carolinensis*, and **Bimodal** (17%), **High** (14%), **Intermediate** (10%) and **Not Responsive** (6%) for *A. picea*.

@@

Despite their similar total numbers of responsive transcripts, the two species differed in the proportions of transcripts in each response category (Table 1). *Aphaenogaster picea* upregulated significantly more genes in response to both temperature extremes (*Bimodal*, ...stats...), while *A. carolinensis* downregulated more transcripts under these conditions (*Intermediate*, ...stats...). Consistent with the *enhanced defense* hypothesis, *A. picea* was more responsive specifically to low temperatures, inducing expression of 1,146 (30%) more transcripts than *A. carolinensis* (Table 1, ... stats...). At high temperatures, *A. carolinensis* showed the opposite pattern, matching the predictions of the *tolerance* hypothesis, inducing expression of ~14% fewer transcripts than *A. picea* (Table 1).

In addition to comparing the overall number of transcripts in each category, we also examined the extent to which reaction norms for individual transcripts were shared between species. Expression patterns differed for 6,792 (65%) of the thermally-responsive transcripts. Most of these shifted from one responsive category to another. These shifts were not randomly distributed (...Stuart-Maxwell test of marginal homogeneity ...) indicating that these differences are not simply due to neutral processes. **Intermediate** transcripts in *A. carolinensis* were significantly more likely to have **Low** or **Bimodal** expression in *A. picea* than expected by chance (...stat...), while there was a significant deficit of the reverse shift of *A. picea* **Intermediate** transcripts to any upregulated category in *A. carolinensis* (Fig. 1). Similarly, both **High** and **Bimodal** transcripts in *A. picea* were more likely to be **Not Responsive** in *A. carolinensis* than expected, while the reverse was not significant (Fig. 1).

The molecular processes in the set of transcripts that switched from down-regulation in *A. carolinensis* to up-regulation in *A. picea* included `r nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea)` GO terms, of which ... were related to metabolism. In addition, there were GO terms related to repair (e.g. "histone modification"), transcript regulation (e.g. "positive regulation of cellular process") and cellular organization (e.g. "transport", "organelle organization"). Altogether, these results suggest that in response to both thermal extremes, the warmer-climate species, *A. carolinensis*, primarily shuts down metabolism while the cooler-climate species, *A. picea*, increases metabolism and initiates damage repair responses.

@@

We tested the *genetic assimilation* hypothesis, a shift from inducible to constitutive expression of stress-response genes, by testing for each temperature extreme whether the suite of transcripts induced in response to that extreme showed a higher constitutive level of expression, here defined as expression at 19.5°C, in the species better adapted to these conditions. Consistent with this hypothesis, we found greater constitutive expression in the warm-climate species *A.carolinensis* than *A. picea* for high temperature responsive transcripts in *A. picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`). However, the reverse pattern of higher constitutive expression in *A. picea* was not observed for low temperature responsive transcripts in *A. carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`).

Instead, the overall pattern is that thermally-responsive transcripts in *A. carolinensis*, regardless of expression pattern, had higher constitutive expression than those in *A. picea*, including those with **Intermediate** expression profiles in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`). An important exception to this pattern is the set of transcripts that had **High** or **Bimodal** expression in *A. picea* but were not thermally responsive in *A. carolinensis* (blue cells in bottom row of Fig. 1). Rather than having constitutively elevated expression as predicted, these genes were uniformally not expressed (...mean TPM...) in *A. carolinensis*.

@@

## *A. carolinensis* tolerates heat stress to a greater extent than *A. picea*

As an additional test of the *tolerance* hypothesis, we examined the timing of gene induction in response to temperature changes. We compared the mean temperatures of transcript upregulation, defined as the temperature at which the transcript showed the greatest positive change in expression, between species. **High** and **Low** transcripts were included only for the direction in which they were upregulated, while **Bimodal** genes were included for analysis at both high and low temperatures. In support of the *tolerance* but not the *enhanced defense* hypothesis, transcripts turned on at high temperatures in *A. carolinensis* were upregulated on average at `r round(T_on.high.ttest$estimate[1], 1)`°C compared to `r round(T_on.high.ttest$estimate[2], 1)`°C for *A. picea* (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 2). For low temperatures, there was no difference in the mean temperature of transcript upregulation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A. picea*; *t*~`r prettyNum(T_on.low.ttest$parameter, digits=2)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 2).

Similarly, for the **Intermediate** transcripts we compared the mean temperatures of down-regulation of transcript expression between species at both high (> 21°) and low (< 21°C) temperatures. Overall, the thresholds for down-regulation were shifted toward cooler temperatures than the corresponding thresholds for gene activation, with a shift of approximately ... at the higher temperature extremes and ...  at low temperature extremes. Again, we found greater inertia of gene expression only in response to high temperature extremes. With increasing temperatures, **Intermediate** transcripts were downregulated at `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A. picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`), while with decreasing temperatures the species did not differ in mean temperature of downregulation (*A. carolinensis* `r round(T_off.int.low.ttest$estimate[1], 1)` compared to *A. picea* `r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`).

@@

The potential for many species to persist in face of contemporary climate change will depend in part upon their thermal tolerance. However, for most species little is known about the plasticity or adaptive changes in gene expression that underlie thermal tolerance. By using the novel *reactionome* approach, we were able to quantitatively describe changes in transcript expression across a thermal gradient. In particular, by characterizing the thermal reactionomes of the ant species *Aphaenogaster picea* and *A. carolinensis* we found patterns consistent with different mechanisms of thermal tolerance in these two species. Overall, we found that about 8% of transcripts (~9,000) are thermally responsive in each species. About 70% of these transcripts have a non-linear response to temperature that differs between species, which would not have been detected by standard differential expression approaches.

Under the *enhanced defense* hypothesis, stress-adapted species are hypothesized to induce a stronger and earlier response to extreme conditions. We found evidence for this hypothesis from the cool-climate species *A. picea* which induced more transcripts in response to low temperatures than did *A. carolinensis* (Table 1, Fig. 1). However, *A. picea* also induced more transcripts in response to high temperatures than did *A. carolinensis*. We also found that higher temperatures were required to induce the expression response for *A. carolinensis* than *A. picea* towards the high thermal limit (Fig. 2B, D), consistent with the *tolerance* but not the *enhanced defense* hypothesis for *A. carolinensis*. There was no difference between species in temperature of response to low temperatures (Fig. 2A,C). ...compare to Drosophila...

>>>>

Thermal tolerance is a key trait governing the distributions of species and is useful in predicting their responses to climate change. To quantitatively evaluate among three hypotheses for the evolution of thermal tolerance, we describe a novel analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient. We characterized the thermal reactionomes of two common ant species in the eastern U.S: the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*. We found that at least 8% of all genes were thermally responsive. Consistent with the *enhanced defense hypothesis*, the cool-adapted *A. picea* induced expression of more genes in response to extreme temperatures than *A. carolinensis*, including heat shock proteins. In contrast, *A. carolinensis* displayed higher constitutive expression of responsive genes and required greater changes towards high, but not low, temperatures to induce changes in gene expression than *A. picea*. This pattern was consistent with the *tolerance hypothesis* at high, but not low, temperatures. We found no evidence for the *genetic assimilation hypothesis* as there was no trade-off between constitutive and inducible levels of gene expression. Collectively, our results indicate that the northern *A. picea* actively resists thermal shock by synthesizing heat shock and other protective molecules to maintain function, whereas the southern *A. carolinensis* relies upon other mechanisms to tolerate thermal stress. These results are consistent with expectations that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.

@@

Coping with progressive changes in climate requires both short-term physiological acclimation and long-term adaptation. To better understand the relationship between these two mechanisms, we compared gene expression as a function of temperature in two closely-related ant species: the cool-climate *Aphaenogaster picea* and the warm-climate *A. carolinensis*. Surprisingly, we found divergent transcriptomic responses in these species. *Aphaenogaster picea* responded to high and low temperatures through upregulation of a larger proportion of its transcriptome with a robust heat shock response. In contrast, *A. carolinensis* required more extreme temperatures to trigger a protective response and downregulated metabolic genes. These results suggest that adaptation to chronic high temperatures involves qualitatively different mechanisms than those used to cope with low temperatures and other stressors.

@@

Temperature regulates biological activity and shapes diversity from molecular to macroecological scales [@brown2004; @kingsolver2009]. Many species, especially small-bodied arthropods, live at temperatures close to their thermal limits and are at risk from current increases in temperature [@deutsch2008; @kingsolver2013; @sunday2014]. Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, depends on a complex interplay between the structural integrity of cellular components and activation of physiological response mechanisms to prevent and/or repair damage [@huey1989; @richter2010]. Thermal defense strategies are shaped by the environmental regime organisms experience [@angilletta_jr2003] and thermal limits vary considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011]. Outside of candidate genes [@krebs1998], little is known about the evolution of thermal tolerance or the link between short-term physiological acclimation and longer-term adaptation to novel temperature regimes. This information is critical for understanding the adaptive potential of species to future climates [@hoffmann2008].

To address this gap of knowledge, we need information on the extent to which selection has acted upon the diversity genes and their plasticity of expression involved in thermal tolerance [@hoffmann2008; @somero2011]. In recent years, whole organism gene expression approaches (e.g. transcriptomics) using high-throughput RNA sequencing (RNAseq) technology have been widely applied to identify genes involved in thermal tolerance [@meyer2011; @teets2012; @barshis2013; @oneil2014] and other traits. Such studies typically use an ANOVA-type experimental or sampling design, with only a few environmental levels, and often find only a few dozen to hundred genes with differential expression in different thermal regimes. However, temperature and other environmental factors vary continuously in nature. As a result, such comparisons (e.g. high vs. low temperatures) are likely to miss key differences that are due not just to whether it is hot, but rather how hot it is. Continuous variation is better characterized with a reaction norm approach, which describes variation in the phenotype of a single genotype across an environmental gradient [@gomulkiewicz1992]. Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014], and differences in the shape of reaction norms are often greater than differences in mean values at both the species and population levels [@murren2014].

Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient. We use this method to identify genes that are thermally responsive in two closely-related eastern North American ant species, *Aphaenogaster carolinensis* (part of the *A. rudis*-species complex, B. DeMarco personal communication) and *A. picea* [@umphrey1996]. *Aphaenogaster* are some of the most common ant in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. *Aphaenogaster carolinensis* (mean annual temperature (MAT) at collection site = 14.6°C, seasonality = 7,678) experiences significantly warmer and less seasonally variable temperatures than *A. picea* (MAT = 4.6°C, seasonality = 10,008; [@hijmans2005]), with correspondingly higher critical maximum temperatures (44.7°C for *A. carolinensis* versus 41.4°C for *A. picea*; Supplemental Material) and minimum temperatures (6.1°C for *A. carolinensis* versus -0.1°C for *A. picea*; Supplemental Material) allowing us to investigate adaptation to both lower and upper thermal extremes in this system.

@@

To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) using Illumina short read sequencing technology, generated the *de novo* transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis (Supplemental Material).

We used these data to quantitatively test three mechanistic hypotheses of thermal adaptation. First, the *enhanced defense hypothesis* [@hofmann1996; @feder1999; @kultz2005] proposes that species extend their thermal limits through a stronger induced responses to provide greater protection from more frequently encountered stressors. This hypothesis would predict that the cool-adapted *A. picea* would induce more genes in response to low temperatures than would the warm-adapted *A. carolinensis*, which would induce more genes in response to high temperatures. Additionally, upregulation of molecular pathways shared by both species should also be stronger for more commonly experienced thermal extremes; low temperatures for *A. picea*, high temperatures for *A. carolinensis*. Second, the *genetic assimiliation hypothesis* [@waddington1953; @sikkink2014] proposes that exposure to more frequent stressors selects for a shift from inducible to constitutive expression of stress-response genes. This hypothesis predicts that transcripts responsive to high temperatures in *A. picea* will have higher constitutive expression in *A. carolinensis*, whereas transcripts responsive to low temperatures in *A. carolinensis* will have higher constitutive expression in *A. picea*. Finally, the *tolerance hypothesis* [@cowles1939; @fields2001] proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural changes [@lockwood2012] or behavioral adapations [@gordon2013; @sunday2014]. This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways. Of course, these hypotheses are not mutually exclusive. By taking a reactionome approach, we can quantify the relative importance of each of these mechanisms to thermal adapatation as the proportion of the transcriptome matching each of these predicted patterns.

@@

The combined *Aphaenogaster* transcriptome assembly contained 98,186 transcripts. About half of these (51,246) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae. Of the Formicidae hits, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts, or 8% of the total, with expression patterns that were thermally responsive at a 5% False Discovery Rate [@benjamini1995]. Of these 9,052 transcripts, 80% (`r prettyNum(length(quadratic.lms), big.mark = ",")`) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g. high vs. low temperature). The proportion of responsive transcripts is the same if we focus only on those transcript with a BLAST hit (4,618 transcripts out of 51,246). However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] suggesting that this estimate of 8% is likely a lower bound for the true proportion of genes that are thermally responsive.

```{r type_table, echo=FALSE, results='asis', fig.cap = "Table of the number of thermally-responsive transcripts by expression type for *A. carolinensis* and *A. picea*. **Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are not thermally responsive in one species but are in the other species."}

@@

The total number of thermally-responsive transcripts did not differ between species (...stat...). Within this set, there was significant qualitative and quantitative variation in thermal responses; both among transcripts and between the two species. We grouped the thermally-responsive transcripts for each species into five expression categories based on their reaction norms: transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (**Not Responsive**). For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C). The majority of the responsive genes were in the **Low** category for both species (42% and 54% for *A. carolinensis* and *A. picea*, respectively), followed by **Intermediate** (29%), **High** (12%),**Bimodal** (10%) and **Not Responsive** (8%) for *A. carolinensis*, and **Bimodal** (17%), **High** (14%), **Intermediate** (10%) and **Not Responsive** (6%) for *A. picea*.

@@

Despite their similar total numbers of responsive transcripts, the two species differed in the proportions of transcripts in each response category (Table 1). *Aphaenogaster picea* upregulated significantly more genes in response to both temperature extremes (*Bimodal*, ...stats...), whereas *A. carolinensis* downregulated more transcripts under these conditions (*Intermediate*, ...stats...). Consistent with the *enhanced defense* hypothesis, *A. picea* was more responsive specifically to low temperatures, inducing expression of 1,146 (30%) more transcripts than *A. carolinensis* (Table 1, ... stats...). In contrast, at high temperatures, *A. carolinensis* matched the predictions of the *tolerance hypothesis*, inducing expression of ~14% fewer transcripts than *A. picea* (Table 1).

In addition to comparing the overall number of transcripts in each category, we also examined the extent to which reaction norms for individual transcripts were shared between species. Expression patterns differed for 6,792 (65%) of the thermally responsive transcripts, most of which shifted from one responsive category to another. These shifts were not randomly distributed (...Stuart-Maxwell test of marginal homogeneity ...) indicating that these differences are not simply due to neutral processes. **Intermediate** transcripts in *A. carolinensis* were significantly more likely to have **Low** or **Bimodal** expression in *A. picea* than expected by chance (...stat...), whereas there was a significant deficit of the reverse shift of *A. picea* **Intermediate** transcripts to any upregulated category in *A. carolinensis* (Fig. 1). Similarly, both **High** and **Bimodal** transcripts in *A. picea* were more likely to be **Not Responsive** in *A. carolinensis* than expected, while the reverse was not significant (Fig. 1).

The molecular processes in the set of transcripts that switched from down-regulation in *A. carolinensis* to up-regulation in *A. picea* included `r nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea)` GO terms, of which ... were related to metabolism. In addition, there were GO terms related to repair (e.g. "histone modification"), transcript regulation (e.g. "positive regulation of cellular process") and cellular organization (e.g. "transport", "organelle organization"). Altogether, these results suggest that in response to both thermal extremes, the warmer-climate species, *A. carolinensis*, primarily shuts down metabolism whereas the cooler-climate species, *A. picea*, increases metabolism and initiates damage repair responses.

@@

We tested the *genetic assimilation* hypothesis, a shift from inducible to constitutive expression of stress-response genes, by examining whether the suite of transcripts induced in response to either temperature extreme showed a higher constitutive level of expression, here defined as expression at 19.5°C, in the species better adapted to these conditions. Consistent with this hypothesis, we found greater constitutive expression in the warm-climate species *A.carolinensis* than *A. picea* for transcripts that were responsive to high temperature in *A. picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`). However, the reverse pattern of higher constitutive expression in *A. picea* was not observed for low temperature responsive transcripts in *A. carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`).

Instead, the overall pattern is that thermally responsive transcripts in *A. carolinensis*, regardless of expression pattern, had higher constitutive expression than those in *A. picea*, including those with **Intermediate** expression profiles in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`). An important exception to this pattern is the set of transcripts that had **High** or **Bimodal** expression in *A. picea* but were not thermally responsive in *A. carolinensis* (blue cells in bottom row of Fig. 1). Rather than having constitutively elevated expression as predicted, these genes were uniformly not expressed (...mean TPM...) in *A. carolinensis*.

@@

## *A. carolinensis* has greater inertia of expression change to increases in temperature than does *A. picea*

As an additional test of the *tolerance hypothesis*, we examined the critical temperature of gene induction in response to increasing and decreasing temperatures. We compared between species the mean temperatures of transcript upregulation, defined as the temperature at which the transcript showed the greatest positive change in expression. In support of the *tolerance hypothesis* but not the *enhanced defense hypothesis*, transcripts turned on at high temperatures were induced on average at `r round(T_on.high.ttest$estimate[1], 1)`°C in *A. carolinensis* compared to `r round(T_on.high.ttest$estimate[2], 1)`°C in *A. picea* (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 2). For low temperatures, there was no difference in the mean temperature of transcript upregulation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A. picea*; *t*~`r prettyNum(T_on.low.ttest$parameter, digits=2)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 2).

Similarly, for the **Intermediate** transcripts we compared the mean temperatures of transcript downregulation of transcript expression between species at both high (> 21°) and low (< 21°C) temperatures. Thresholds for downregulation were shifted toward cooler temperatures than were the corresponding thresholds for gene activation, with a shift of approximately ... at the higher temperature extremes and a shift of ...  at low temperature extremes. Again, there was greater inertia of gene expression only in response to high temperature extremes. With increasing temperatures, **Intermediate** transcripts were downregulated at `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A. picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`), whereas with decreasing temperatures the species did not differ in mean temperature of downregulation (*A. carolinensis* `r round(T_off.int.low.ttest$estimate[1], 1)` compared to *A. picea* `r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`).

@@

Collectively, our results indicate that these two species have distinct mechanisms of response to thermal extremes. The cool-adapated species *Aphaenogaster picea* mounts a classical cellular stress response [@kultz2005; @richter2010] with elevated gene expression at both low and high temperatures, which has been shown to impose a metabolic cost [@hoekstra2013]. In contrast, the warm-adapted *A. carolinensis* relies upon constitutive defenses [@barshis2013], and perhaps enhanced proteome stability [@ghosh2010] or behavioral responses [@sunday2014] to tolerate thermal stress. If the *enhanced defense* is hypothesis is correct, candidate genes such as heat shock proteins should be over-represented in the thermally-responsive category for *A. picea*. Of the `r nrow(hsp_all)` transcripts annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to thermal extremes in *A. picea* whereas only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A. carolinensis*. These differences are in line with the expectations if *A. picea*, a northern species with fewer good foraging days, must forage in suboptimal conditions whereas *A. carolinensis* can, with a longer season, afford to restrain from foraging in suboptimal conditions. In fact, restrained foraging under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].
